ODGI: understanding pangenome graphs

### 2021-10-29 manuscript/main.tex
odgi build performance figure
update fig letter in odgi viz/draw section

We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA. Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes. Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (TODO).

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
White spaces indicate where paths do not traverse the nodes.
The meaning of the colors depends on how \textit{odgi viz} is executed.
By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{a}).
Path names are displayed on the left of the paths.
The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig. \ref{fig:odgi_viz}\textbf{b-d}), making it easy to grasp the properties of the graph and its complexity.
Graph nodes order can affects downstream analyses on pangenome graphs. With \textit{odgi viz} we can color the paths by path position (Fig. \ref{fig:odgi_viz}\textbf{b}), with light grey indicating where paths begin and dark grey where they ends.
This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
Pangenome graphs represents both strand of DNA sequences.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig. \ref{fig:odgi_viz}\textbf{b}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig. \ref{fig:odgi_viz}\textbf{d}).

\textit{odgi draw} extends the visualization in 2 dimensions (2D) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}). A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.

@@

>>>>

We begin by transforming the storage model of the GFA format (in which nodes, edges, and paths are described independently) into the ODGI node-centric encoding with \textit{odgi build}.
This construction step can present a significant bottleneck, in particular as the size of the path set of the graph increases.
The ODGI data structure (Algorithm \ref{alg:structs}) allows algorithms that build and modify the graph to operate in parallel, without any global locks.
In \textit{odgi build}, we initially construct the node vector in a serial operation that scans across the input GFA. Then, in serial, we add edges in the $Node.edges$ vectors of pairs of nodes. Finally, we create paths in serial, and extend them in parallel by obtaining mutex $Node.lock$ for pairs of nodes and by adding the path step in their $Node.path\_steps$ vectors.
This parallelism speeds ODGI model construction by many-fold when testing against graphs built by the HPRC (Figure \ref{fig:build}).

@@

\textit{odgi viz} supports a binned, linearized rendering in 1 dimension (that is, all graph nodes lie on the same axis).
This visualization is computed in linear-time and offers a human-interpretable format suitable for understanding the topology and genome relationships in the pangenome graph (Fig. \ref{fig:odgi_viz}).
Graph nodes are arranged on a single axis, from left to right, with the colored bar indicating the paths and the nodes they cross.
White spaces indicate where paths do not traverse the nodes.
The meaning of the colors depends on how \textit{odgi viz} is executed.
By default, path colors has no particular meaning, but each path has its own color (Fig. \ref{fig:odgi_viz}\textbf{b}).
Path names are displayed on the left of the paths.
The black lines on the bottom indicate the edges connecting the nodes and, therefore, represent the graph topology.

Nevertheless, complex, nonlinear graph structures are difficult to display and interpret in a low number of dimensions.
To overcome such a limitation, \textit{odgi viz} supports multiple visualization modalities (Fig. \ref{fig:odgi_viz}\textbf{c-e}), making it easy to grasp the properties of the graph and its complexity.
Graph nodes order can affects downstream analyses on pangenome graphs. With \textit{odgi viz} we can color the paths by path position (Fig. \ref{fig:odgi_viz}\textbf{c}), with light grey indicating where paths begin and dark grey where they ends.
This visualization is suitable for checking graph nodes' linear order, as smooth color gradients indicate that the graph reflects well the linear paths' coordinate system.
Pangenome graphs represents both strand of DNA sequences.
\textit{odgi viz} supports also coloring the paths by orientation, with paths colored in red and black where their sequence is reversed or in forward, respectively, with respect to the sequences represented by the graph nodes (Fig. \ref{fig:odgi_viz}\textbf{d}).
Eukaryotic genomes experience gains and losses of genetic material, resulting in copy number variation (CNV) across the population.
With \textit{odgi viz}, we can use multiple color palettes to color the paths by path depth, highlighting the different copy number statuses in the genomes represented in the pangenome graph (Fig. \ref{fig:odgi_viz}\textbf{e}).

\textit{odgi draw} extends the visualization in 2 dimensions (2D)  (Fig. \ref{fig:odgi_viz}\textbf{a}) by rendering the layout built by \textit{odgi layout} (\S\ref{sec:sort}). A 2D rendering is more costly to compute, but we similarly provide a linear-time scaling implementation, allowing us to apply it to large pangenome graphs.

@@

\input{fig_build_time}
